---
title: "Assignment 1"
author: "Colin Jones (100849896)"
date: "2023-01-11"
output:
  html_document:
    css: "style.css"
    toc: true
    toc_float:
      toc_collapsed: true
    toc_depth: 3
    number_sections: false
    theme: lumen
---

<script>
   $(document).ready(function() {
     $head = $('#header');
     $head.prepend('<img src=\"logo.jpg\" style=\"float: right;width: 200px;\"/>')
   });
</script>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The version of `R` is as listed below.

```{r}
version
```

The version of `Python` is listed below.

```{python}
import sys
sys.version
```

## Objective

The purpose of this Assignment is to recreate the base `R` code of the textbook in higher-level `R` and `Python`. There will be three different blocks of code that do the same thing: one in base `R`, one in fancy `R`, and one in `Python`. The original base `R` code is taken from <https://hastie.su.domains/ISLR2/Labs/Rmarkdown_Notebooks/Ch2-statlearn-lab.html>. All other code is my own.

## Libraries

First we install the packages needed for this notebook (note that I already have done it so it is commented out). Then we load the libraries.

```{r error=FALSE, warning=FALSE, message=FALSE}
# install.packages("dplyr", "plotly", "htmlwidgets", "GGally")
library(dplyr) # Used for fancier R manipulation
library(plotly) # Used for fancier plots
library(htmlwidgets) # Used to save plotly plots
library(GGally) # Used for pairs plots
```

Next for `Python` we install the packages with `pip`. Again I already have done that in the terminal. You can run it right here with the `%` in front of it (But this is a little hacky).

```{python}
# %pip install plotly
# %pip install pandas
# %pip install numpy
```

And we load them.

```{python}
import numpy as np # For vectors, matrices etc.
import plotly.express as px # For plotting simple graphs
import plotly.graph_objects as go # For plotting more complex graphs
import pandas as pd # For data frames
```

## Basic Commands

### Vectors

The first block of code assigns a vector. Note that the older assignment in `R`, `<-`, is just `=` in `Python`:

```{r}
# In R
x <- c(1, 3, 2, 5)
x
```

In `Python` we use `[]` to create a list. In order to create a vector we need to use the linear algebra library `numpy`.

```{python}
# In Python
x = np.array([1,3,2,5])
x
```

Similarly for the next vector with just `=` in `R`.

```{r}
# In R
x = c(1, 6, 2)
x
```

```{python}
# In Python
x = np.array([1, 6, 2])
x
```

```{r}
# In R
y = c(1, 4, 3)
```

```{python}
# In Python
y = np.array([1, 4, 3])
```

Next is the `length()` function which for numpy arrays is `shape`.

```{r}
# In R
length(x)
length(y)
```

And `Python`.

```{python}
# In Python
x.shape
y.shape
```

And the summation.

```{r}
# In R
x + y
```

```{python}
# In Python
x + y
```

### Variables

Next we look at the `ls()` function. In `Python` it can be done with the `dir()` function. `Python` gives a bit more info with some built-in things and packages.

```{r}
# In R
ls()
```

```{python}
# In Python
dir()
```

Now we remove some variables.

```{r}
# In R
rm(x, y)
ls()

```

```{python}
# In Python
del x,y
dir()
```

And all objects at once.

```{r}
# In R
rm(list = ls())
```

To delete only user-defined variables in `Python` we only want ones that don't start with `__`.

```{python}
# In Python
for obj in dir():
  if not obj.startswith("__"):
    del globals()[obj]
dir()

```

### Asking for help

Next we ask for help with the matrix function. (Uncomment if you choose; it opens a pop up.)

```{r}
# In R
# ?matrix

```

```{python}
# In Python
import numpy as np # Import again because we just removed it
# help(np.array)

```

### Matrices (or arrays if you are very dimensional)

Now we build some matrices.

```{r}
# In R
x <- matrix(data = c(1, 2, 3, 4), nrow = 2, ncol = 2)
x

```

For `Python` it is a little bit different. We can pass a single vector but then we have to reshape it to be a matrix.

```{python}
# In Python
x = np.array([1,2,3,4]).reshape(2,2)
x
```

If you notice `Python` reshaped it by filling rows first. To get the matrix we want we need to transpose it.

```{python}
# In Python
x = np.array([1,2,3,4]).reshape(2,2) \
                       .transpose()
x

```

This can be done more easily in `R` with the `byrow` flag.

```{r}
# In R
x = matrix(c(1, 2, 3, 4), 2, 2, byrow = TRUE)

```

Next we square root and square the matrices.

```{r}
# In R
sqrt(x)
x^2
```

And `Python`.

```{python}
# In Python
np.sqrt(x)
np.square(x)
```

### Random numbers

Finally we generate some random numbers. In `Python` we need to specify $\mu$ and $\sigma$ whereas in `R` it is by default 0 and 1. `Python` produces a correlation matrix instead. Also the numbers are different due to different seeds.

```{r}
# In R
x <- rnorm(50)
y <- x + rnorm(50, mean = 50, sd = .1)
cor(x, y)
```

```{python}
# In Python
x = np.random.normal(0, 1, 50)
y = x + np.random.normal(50, .1, 50)
np.corrcoef(x,y)
```

We can set the seed to make reproducible code.

```{r}
# In R
set.seed(3)
y <- rnorm(100)
mean(y)
```

```{python}
# In Python
np.random.seed(3)
y = np.random.normal(0, 1, 100)
np.mean(y)
```

Finally we ask for some variances and standard deviations.

```{r}
# In R
var(y)
sqrt(var(y))
sd(y)
```

Repeat in `Python`.

```{python}
# In Python
np.var(y)
np.sqrt(np.var(y))
np.std(y)
```

## Graphics

`plotly` is a java-based interactive plotting library. It shares many similarities with `ggplot` in `R`. In `Python` there are a bit more nuances, mostly if you want it simple or complicated.

```{r}
x <- rnorm(100)
y <- rnorm(100)
plot(x, y)
```

Next in fancy `R`.

```{r}
# In Fancy R
data.frame(cbind(x,y)) |>
  plot_ly(x=x, y=y) |>
  add_markers() |>
  layout(title="Plot of Y vs X", xaxis=list(title="this is the x-axis"),
         yaxis=list(title="this is the y-axis"))
```

And `Python`

```{python}
# In Python
import plotly.express as px # Again we got rid of it too with the delete all
fig = px.scatter(x=np.random.normal(0,1,100), y=np.random.normal(0,1,100),
                 title="Plot of X vs Y", labels=dict(x="this is the x-axis",
                                                     y="this is the y-axis"))
fig.show()
```

Now to save it.

```{r}
# In R
library(htmlwidgets) # Back again

p <- data.frame(cbind(x,y)) |>
  plot_ly(x=x, y=y) |>
  add_markers() |>
  layout(title="Plot of X vs Y", xaxis=list(title="this is the x-axis"),
         yaxis=list(title="this is the y-axis"))

saveWidget(p, file="scatter.html")

```

And `Python`.

```{python}
# In Python
fig = px.scatter(x=np.random.normal(0,1,100), y=np.random.normal(0,1,100),
                 title="Plot of X vs Y", labels=dict(x="this is the x-axis",
                                                     y="this is the y-axis"))

fig.write_html("scatter2.html")

```

### Sequences and Surfaces

Next we generate some sequences in `R` and `Python` using the `range()` function. Don't forget `Python` indexes from 0. We also have to use the `np.linspace()` function in `Python`.

```{r}
# In R
x <- seq(1, 10)
x <- 1:10
x <- seq(-pi, pi, length = 50)
```

`Python`:

```{python}
# In Python
x = range(1,11)
x = np.linspace(-np.pi, np.pi, 50)
```

Now that we have a domain, we can define a function and plot it as a contour map as well as a surface.

```{r}
# In R
y <- x
f <- outer(x, y, function(x, y) cos(y) / (1 + x^2))
contour(x, y, f)
contour(x, y, f, nlevels = 45, add = T)
```

I chose a different number of levels due to data visualization standards. The base `R` way is just too cluttered.

```{r}
# Fancy R
as.data.frame(cbind(x,y,f)) |>
  plot_ly(x=x, y=y) |>
    add_contour(z=matrix(f, nrow = length(y), byrow = TRUE),
                contours = list(
                    start=-.8,
                    end=.8,
                    size=.1,
                    showlabels = TRUE,
                    coloring="lines")) |>
    layout(title="Contour plot", xaxis=list(title="x"), yaxis=list(title="y"))
```

For `Python` things are similar but we use the `meshgrid` function.

```{python}
# In Python
y = x
xr,yr = np.meshgrid(x,y) # A way of making a continuous domain of R^2
f = np.cos(yr) / (1 + xr**2)
```

```{python}
# In Python
import plotly.graph_objects as go # We really shouldn't have deleted everything ...

fig = go.Figure() \
        .add_contour(x=x,y=y,z=f, contours=dict(
            start=-.8,
            end=.8,
            size=.1,
            showlabels=True,
            coloring="lines")) \
        .update_layout(title="Contour plot") \
        .update_xaxes(title="x") \
        .update_yaxes(title="y")
fig.show()

```

Another surface.

```{r}
# In R
fa <- (f - t(f)) / 2

contour(x, y, fa, nlevels = 15)

```

```{r}
# Fancy R
as.data.frame(cbind(x,y,fa)) |>
  plot_ly(x=x, y=y) |>
    add_contour(z=matrix(fa, nrow = length(y), byrow = TRUE),
                contours = list(
                    start=-.8,
                    end=.8,
                    size=.1,
                    showlabels = TRUE,
                    coloring="lines")) |>
    layout(title="Contour plot 2", xaxis=list(title="x"), yaxis=list(title="y"))

```

And `Python`.

```{python}
# In Python
fa = (f - f.T) / 2

fig = go.Figure() \
        .add_contour(x=x,y=y,z=fa, contours=dict(
            start=-.8,
            end=.8,
            size=.1,
            showlabels=True,
            coloring="lines")) \
        .update_layout(title="Contour plot") \
        .update_xaxes(title="x") \
        .update_yaxes(title="y")

fig.show()

```

Let's fill things in.

```{r}
# Fancy R
as.data.frame(cbind(x,y,fa)) |>
  plot_ly(x=x, y=y) |>
    add_contour(z=matrix(fa, nrow = length(y), byrow = TRUE),
                contours = list(
                    start=-.8,
                    end=.8,
                    size=.1,
                    showlabels = TRUE,
                    labelfont=list(color="black"))) |>
    layout(title="Image plot", xaxis=list(title="x"), yaxis=list(title="y"))

```

```{python}

# In python
fig = go.Figure() \
        .add_contour(x=x,y=y,z=fa, contours=dict(
            start=-.8,
            end=.8,
            size=.1,
            showlabels=True,
            labelfont=dict(color="black"))) \
        .update_layout(title="Image plot") \
        .update_xaxes(title="x") \
        .update_yaxes(title="y")
fig.show()

```

Going 3D and adding some perspective.

```{r}
# Fancy R
as.data.frame(cbind(x,y,fa)) |>
  plot_ly(x=x, y=y) |>
    add_surface(z=matrix(fa, nrow = length(y), byrow = TRUE)) |>
    layout(title="Perspective plot", xaxis=list(title="x"), yaxis=list(title="y"))

```

```{python}
# In python
fig = go.Figure() \
        .add_surface(x=x,y=y,z=fa) \
        .update_layout(title="Perspective plot") \
        .update_xaxes(title="x") \
        .update_yaxes(title="y")

fig.show()

```

## Indexing Data

Now we slice and dice the arrays. Depending on who you ask, remember what you consider to start the natural numbers. First define `A`.

```{r}
# In R
A <- matrix(1:16, 4, 4)
A
```

`Python`:

```{python}
# In Python
A = np.array(range(1,17)).reshape(4,4).transpose()
A

```

Slice away.

```{r}
# In R
A[2, 3]
```

```{python}
# In Python
A[1,2]
```

```{r}
# In R
A[c(1, 3), c(2, 4)]
```

Not so easy in `Python`. Took a while to find the `ix_` function.

```{python}
# In Python
A[(np.ix_([0,2], [1,3]))]
```

```{r}
# In R
A[1:3, 2:4]
```

```{python}
# In Python
A[:3, 1:4]
```

```{r}
# In R
A[1:2, ]
```

```{python}
# In Python
A[:2,:]
```

```{r}
# In R
A[, 1:2]

```

```{python}
# In Python
A[:, :2]
```

```{r}
# In R
A[1, ]
```

```{python}
# In Python
A[0,:]
```

```{r}
# In R
A[-c(1, 3), ]
```

```{python}
# In Python
np.delete(A, [0,2], 0)
```

```{r}
# In R
A[-c(1, 3), -c(1,3,4)]
```

```{python}
# In Python
np.delete(np.delete(A, [0,2], 0), [0,2,3], 1).flatten()
```

```{r}
# In R
dim(A)
```

```{python}
# In Python
A.shape
```

## Loading Data

Now let's get to some data. We read in a data set from the website https://hastie.su.domains/ISLR2/Labs/. It is about cars. Read in and take a look.

```{r}
# In R
Auto <- read.csv("Auto.csv", header = T, na.strings = "?", stringsAsFactors = T)
View(Auto)
head(Auto)
dim(Auto)
```

```{python}
# In Python
import pandas as pd

Auto = pd.read_csv("Auto.csv")
Auto.head()
Auto.shape

```

```{r}
# In R
Auto[1:4, ]
```

Now `dplyr` comes into the picture.

```{r}
# Fancy R
Auto |> slice(1:4)
```

```{python}
# In Python
Auto.iloc[:4,:]
```

```{r}
# In R
Auto <- na.omit(Auto)
dim(Auto)
```

```{r}
# Fancy R
Auto |> na.omit() |>
        dim()

```

```{python}
# In Python
Auto.dropna().shape
```

```{r}
# In R
names(Auto)
```

```{r}
# Fancy R
Auto |> names()
```

```{python}
# In Python
Auto.columns
```

## Additional Graphical and Numerical Summaries

Now let's do some more plotting.

```{r}
# In R
plot(Auto$cylinders, Auto$mpg)
```

```{r}
# Fancy R
Auto |> plot_ly(x=~cylinders, y=~mpg) |>
        add_markers() |>
        layout(title="Number of Cylinders vs Miles per Gallon",
               xaxis=list(title="Cylinders"),
               yaxis=list(title="MPG"))
```

```{python}
# In Python

fig = px.scatter(Auto, x="cylinders", y="mpg",
                 title="Number of Cylinders vs Miles per Gallon",
                 labels=dict(cylinders="Cylinders",
                 mpg="MPG"))

fig.show()

```

### Box plots

You can `attach()` things if you are so inclined that just makes the variable names available in `R`. The plot remains the same. The document from the textbook website plots many box plots changing one feature at a time. I have put them all together so you don't see so many. Note that `plotly` does not have the `varwidth` option. Instead I added jitter points so you can see how many observations there are.

```{r}
# In R
attach(Auto)

cylinders <- as.factor(cylinders)
plot(cylinders, mpg, col = "red", varwidth = T)

```

```{r}
# Fancy R
Auto |> mutate(cylinders = as.factor(cylinders)) |>
        plot_ly(x=~mpg, y=~cylinders, color="red") |>
        add_boxplot(line=list(color="red"),marker=list(color="red"), boxpoints="all", jitter=.3 ) |>
        layout(title="Number of Cylinders vs Miles per Gallon",
               xaxis=list(title="MPG"),
               yaxis=list(title="Cylinders"))

```

```{python}
# In Python

Auto['cylinders'] = Auto['cylinders'].astype("object")

fig = px.box(Auto, x="mpg", y="cylinders", points="all",
             title="Number of Cylinders vs Miles per Gallon",
             labels=dict(cylinders="Cylinders", mpg="MPG"))

fig.show()

```

### Histograms

And now for some histograms. Again they plot many of them; I will just plot one version with all the features in it.

```{r}
# In R
hist(mpg, col = 2, breaks = 15)
```

```{r}
# Fancy R
Auto |> plot_ly(x=mpg) |>
        add_histogram(xbins=15, color="red", stroke=list(color="black")) |>
        layout(title="Histogram of MPG",
               xaxis=list(title="MPG"),
               yaxis=list(title="Frequency"))
```

```{python}
# In Python
fig = px.histogram(Auto, x="mpg",
             title="Number of Cylinders vs Miles per Gallon",
             labels=dict(mpg="MPG"))

fig.show()

```

### Pairs plot

Finally the pairs plot. Let's not plot them all like they do in the lab because it is too much and overwhelming.

```{r}
# In R
pairs(
    ~ mpg + displacement + horsepower + weight + acceleration,
    data = Auto

  )

```

And now for an upgrade.

```{r fig.height=8, fig.width=12}
# Fancy R
(ggpairs(Auto, columns = c(1,3:6), title="Pairs Plot")) |> ggplotly()
```

```{python fig.height=10, fig.width=14}
# In Python
fig = px.scatter_matrix(Auto[["mpg","displacement","horsepower","weight", "acceleration"]],
                        title="Pairs Plot", width=1000, height=1000)

fig.show()

```

### Summaries

For the last part we make summaries of the data frames and variables. For the whole frame,

```{r}
# In R
summary(Auto)

```

```{r}
# Fancy R
Auto |> summary()

```

```{python}
# In Python
Auto.describe()

```

And with one variable.

```{r}
# In R
summary(mpg)

```

```{r}
# Fancy R
Auto |> select(mpg) |>
        summary()

```

```{python}
# In Python
Auto["mpg"].describe()

```
